The Concurrent Calculi Formalisation Benchmark

### 2023-10-09 paper/paper.tex
Francisco: a pass over the introduction

The introduction of the influential POPLMark challenge~\cite{POPLMark}
spearheaded a shift to including mechanised proofs accompanying
published papers. POPLMark's abstract starts with an important
question: ``How close are we to a world where every paper on programming languages is accompanied by an electronic appendix with machine-
checked proofs?''. Much progress has been done in that respect,
notably the POPLMark Reloaded~\cite{POPLMarkReloaded} challenge aims
to identify and develop best practices and tool support for proofs
using logical relations.

The authors of both POPLMark and POPLMark Reloaded
were careful to note that their benchmarks were only a beginning, and
they specifically mention reasoning about concurrency using
coinduction and linear environments as points for future work. Papers
in concurrency theory do often include mechanised proofs
(e.g.~\cite{DBLP:conf/pldi/Castro-Perez0GY21,DBLP:conf/tacas/CastroFY20,lmcs:9985,
  DBLP:journals/jar/CruzFilipeMP23, Tirore:2023}), and others also
directly discuss formalisation techniques
(e.g.~\cite{DBLP:journals/jar/BengtsonPW16, DBLP:conf/tphol/Gay01,
  DBLP:conf/ppdp/Thiemann19, DBLP:conf/forte/ZalakainD21}). However,
our experience is that choosing techniques and tools remains a
challenge with few guidelines. We propose a new collection of
benchmark problems specifically designed to tackle the issues
encountered when mechanising message-passing concurrency using process
calculi. A goal of our collection of benchmark problems is to push for
the development of guidelines and tutorials, and our challenge
problems are a simple setting in which to demonstrate and compare
techniques. Another goal is to strengthen the culture of mechanisation
by rallying the community to collaborate on exploring and developing
new tools and techniques. We have identified three key aspects which
cause issues when mechanising concurrency theory: linearity, scope
extrusion in name passing systems and infinite behaviour modelled
coinductively. Each part of our benchmark exercises one of the three aspects
independently to enable comparison and to keep the challenge problems
small. The three aspects are of course not the only ones that may
cause issues in mechanisations, but they are fundamental to the field
and will thus appear in most mechanisations of concurrency theory.
Mechanisations of actual research will additionally often require more than just
one aspect, but we do not address the combination of techniques.

Large scale formalisation of concurrency theory is already possible.
(Paragraphs about related work in this vein)

However, it is our experience that tools and techniques are chosen in a very ad hoc fashion.
Like previous benchmark authors, we thus seek to answer a number of questions:
\begin{enumerate}
\item What is the current state of the art in mechanising process calculi?
\item What techniques and best practices can be recommended when starting formalisation projects within concurrency theory?
\item What improvements are needed to make mechanisation tools more user-friendly with regards to the issues faced when mechanising process calculi?
\end{enumerate}

Our challenge problems are designed to stimulate progress on answering these questions by providing simple environments in which to demonstrate and evaluate the many tools and techniques available.
Our problems are drawn from the basic metatheory of process calculi and include as little syntax and as few definitions as possible to obtain an environment in which to express theorems that require interesting proof techniques.
The theorems themselves are all well-known results, which suits our aim of exploring mechanisation techniques instead of new theory.
The small scale of the challenge problems also means that they are also accessible to mechanisation experts who may not be familiar with concurrency theory.
While the challenge problems themselves are thus not of scientific interest, we expect that the best practices and mechanisations developed by solving the challenge problems will also be useful for future research.

>>>>

The introduction of the influential POPLMark challenge~\cite{POPLMark}
spearheaded a shift to publications that include mechanised proofs.
POPLMark's abstract starts with an important question: ``How close are
we to a world where every paper on programming languages is
accompanied by an electronic appendix with machine- checked proofs?''.
Much progress has been done in that respect, proof assistants,
libraries, and best practices evolved from this challenge.
Furthermore, extensions such as the POPLMark
Reloaded~\cite{POPLMarkReloaded} challenge aim to identify and develop
best practices and tool support for proofs using logical relations.

The authors of both POPLMark and POPLMark Reloaded both explicitly
note that their benchmarks were only a beginning. Furthermore, they
specifically mention reasoning about concurrency using coinduction and
linear environments as points for future work. While papers in
concurrency theory do often include mechanised proofs
(e.g.~\cite{DBLP:conf/pldi/Castro-Perez0GY21,DBLP:conf/tacas/CastroFY20,lmcs:9985,
  DBLP:journals/jar/CruzFilipeMP23, Tirore:2023}), and others also
directly discuss formalisation techniques
(e.g.~\cite{DBLP:journals/jar/BengtsonPW16, DBLP:conf/tphol/Gay01,
  DBLP:conf/ppdp/Thiemann19, DBLP:conf/forte/ZalakainD21}), our
experience is that choosing techniques and tools remains a challenge
with few guidelines. Thus, in order to further increase the reach of
POPLMark challenge, we propose a new collection of benchmark problems
specifically designed to tackle the issues encountered when
mechanising message-passing concurrency using process calculi.

A goal of our collection of benchmark problems is to push for the
development of guidelines and tutorials, and our challenge problems
are a simple setting in which to demonstrate and compare techniques,
libraries, and proof assistant features. Another goal is to strengthen
the culture of mechanisation by rallying the community to collaborate
on exploring and developing new tools and techniques. We have
identified three key aspects which cause issues when mechanising
concurrency theory: linearity, scope extrusion in name passing systems
and infinite behaviour modelled coinductively. Each part of our
benchmark exercises one of the three aspects independently to enable
comparison and to keep the challenge problems small. The three aspects
are of course not the only ones that may cause issues in
mechanisations, but they are fundamental to the field and will thus
appear in most mechanisations of concurrency theory. Mechanisations of
actual research will additionally often require more than just one
aspect, but we do not address the combination of techniques.

Large scale formalisation of concurrency theory is already possible.
%(Paragraphs about related work in this vein)
However, it is our experience that tools and techniques are chosen in
a very ad hoc fashion. Moreover, it is our opinion that the community
will benefit from a set of solutions exploring different ways of
mechanising typical problems in the field. Like previous benchmark
authors, we thus seek to answer a number of questions:
\begin{enumerate}
\item What is the current state of the art in mechanising process calculi?
\item What techniques and best practices can be recommended when starting formalisation projects within concurrency theory?
\item What improvements are needed to make mechanisation tools more user-friendly with regards to the issues faced when mechanising process calculi?
\end{enumerate}

Our challenge problems are designed to stimulate progress on answering these questions by providing simple environments in which to demonstrate, evaluate, and compare the many tools and techniques available.
Our problems are drawn from the basic metatheory of process calculi and include as little syntax and as few definitions as possible to obtain an environment in which to express theorems that require interesting proof techniques.
The theorems themselves are all well-known results, which suits our aim of exploring mechanisation techniques instead of new theory.
The small scale of the challenge problems also means that they are also accessible to mechanisation experts who may not be familiar with concurrency theory.
While the challenge problems themselves are thus not of scientific interest, we expect that the best practices and mechanisations developed by solving the challenge problems will also be useful for future research.
